Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-03-20 transmission.tex
make true vs posterior plot clearer
move half of plot to appendix

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for (A) all of England and (B) its component regions.
        Yorkshire and the Humber is labelled Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
    \label{transmission:fig:backcalc-alpha}
\end{figure}

@@

For $\vec{\pi}$ and $\rho$, the correlation between the posterior median and the generative value is weak (see \cref{SEIR:fig:true-vs-posterior}).
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
    \todo[inline]{Fix x-axis labels overlapping / being unreadable}
\end{figure}

@@

Overall, the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
\todo{move one or some of the goodness-of-fit plots into an appendix - maybe by region?}
There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
        The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev-young}
\end{figure}
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
    }
    \label{SEIR:fig:prev-old}
    \todo[inline]{Remake/reposition this plot so that it fits on a page.}
\end{figure}

@@

The current model could be more flexible with regards to the age distribution of infections.
This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
This implies that the data would allow identification of a more flexible age-based model.

@@

Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
This is because the model is not flexible enough to capture the true age distribution of infections.
The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}

@@

\ifSubfilesClassLoaded{
  \appendix
  \subfile{transmission-appendix-INLA}
  \subfile{transmission-appendix-phase-type}
  \listoftodos
}{}

>>>>

During November, incidence starts rising, with regional heterogeneity in the timing of when this occurs (see \cref{transmission:fig:backcalc-regions}).
The most explored explanation for this is the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country~\autocite{walkerTracking,daviesEstimated,lythgoeLineage}.
There is evidence that when incidence starts increasing is correlated with when the Alpha variant's spread (see \cref{transmission:fig:backcalc-alpha}).
A second peak occurs in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for (A) all of England and (B) its component regions.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}
\begin{figure}
    \includegraphics{transmission/backcalc-alpha}
    \caption[Correlation of the emergence of Alpha and minimum incidence.]{%
        x-axis: time that S-gene target failure, a marker of the Alpha variant, is estimated to start increasing from baseline by region among CIS test results; from \textcite[table S3]{walkerTracking}.
        y-axis: posterior median and 95\% CrI of the date of minimum incidence in November or early December in each region.
        Linear regression line shown.
    }
    \label{transmission:fig:backcalc-alpha}
\end{figure}

@@

For $\vec{\pi}$ and $\rho$, the correlation between the posterior median and the generative value is weak (see \cref{SEIR:fig:true-vs-posterior}).
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
\end{figure}

@@

Overall, the model fits the data well (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_main}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit]{%
        Goodness-of-fit to prevalence data by region and age for selected regions (see \cref{SEIR:fig:prev-appendix} for others).
        The central ribbon and line show the posterior median and 95\% CrI estimate for the proportion of the relevant stratum RT-PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev}
\end{figure}

@@

The current model could be more flexible with regards to the age distribution of infections.
This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
In some regions, there appears to be a trade-off between under and overfitting in different age groups (see \cref{SEIR:fig:prev}).
For example, in the North East, most of the data points in the $[11, 16)$ age group are below the median posterior estimate, while most of the data points in the $[16, 25)$ age group are above the median posterior estimate.
This implies that the data would allow identification of a more flexible age-based model.

@@

Many of these issues are clearly demonstrated by the estimates for the North East (see \cref{transmission:fig:compare-NE}).
I previously observed the quality of the SEIR model's fit is variable between different age groups (see \cref{SEIR:fig:prev,SEIR:fig:prev-appendix}).
This is because the model is not flexible enough to capture the true age distribution of infections.
The overestimation of prevalence in the 11--16 age group and underestimation in the 16--25 age group is reflected when comparing incidence estimates to those from the backcalculation model, which have reduced bias.
\begin{figure}
    \centering \includegraphics{transmission/compare-NE}
    \caption[Comparing each models estimate of North East incidence by age.]{%
        Incidence proportion in each age group in the North East estimated by the two models discussed in this chapter.
        Posterior median and 95\% CrI shown.
        For the SEIR model, newly RT-PCR positive (entrances into the P1 compartment), rather than incidence (entrances into the E compartment) are shown for comparability with the backcalculation results.
    }
    \label{transmission:fig:compare-NE}
\end{figure}

@@

\ifSubfilesClassLoaded{
  \appendix
  \subfile{transmission-appendix-INLA}
  \subfile{transmission-appendix-phase-type}
  \subfile{transmission-appendix-plots}
  \listoftodos
}{}
